---
title: "BIOL 432 — Assignment 9: Floristic Survey and NMDS"
author: "Justin Lasrado"
date: "2025-10-28"
output: html_document
---

## Overview

This assignment looks at plant community composition and species richness inside and outside garlic mustard plots. I used NMDS to visualize community composition and an ANOVA to look at the effects of Location and Population on the first NMDS axis.

```{r setup, echo=TRUE, message=FALSE}
library(vegan)

floristic <- read.csv("FloristicSurvey.csv")
head(floristic)
str(floristic)

# Separate the metadata from the species data
meta <- floristic[, 1:10]
species <- floristic[, 11:ncol(floristic)]

# Scale each species by its maximum value
species_scaled <- species
for (i in 1:ncol(species_scaled)) {
  species_scaled[, i] <- species_scaled[, i] / max(species_scaled[, i])
}
```

## NMDS

```{r nmds}
# Calculate Bray-Curtis distances and run the NMDS
dist_matrix <- vegdist(species_scaled, method = "bray")
nmds <- metaMDS(dist_matrix, k = 2)
nmds
```

```{r nmds-plot}
# Plot the sites and colour them by location
plot(nmds, display = "sites", type = "n", main = "NMDS of Plant Communities")
points(
  nmds,
  display = "sites",
  col = ifelse(meta$Location == "i", "red", "blue"),
  pch = 19
)
legend("topright", legend = c("Inside", "Outside"), col = c("red", "blue"), pch = 19)
```

## Species richness

```{r richness}
# Count the number of species present at each site
meta$Richness <- rowSums(species > 0)

boxplot(
  Richness ~ Location,
  data = meta,
  main = "Species Richness by Garlic Mustard Presence",
  col = c("blue", "red"),
  names = c("Outside", "Inside"),
  ylab = "Number of Species"
)
```

## ANOVA

```{r anova}
# Add the NMDS coordinates to the metadata
site_scores <- scores(nmds, display = "sites")
meta$NMDS1 <- site_scores[, 1]
meta$NMDS2 <- site_scores[, 2]

# Test the effects of Location and Population on NMDS1
anova_both <- aov(NMDS1 ~ Location + Population, data = meta)
summary(anova_both)
```

## Results

The NMDS plot shows that the Inside and Outside sites are mixed rather than forming two clearly separated groups. The species richness plot also shows similar values between the two locations.

The ANOVA found a strong effect of Population on NMDS1 (F = 49.59, p = 1.43e-07), while Location was not significant (F = 0.35, p = 0.562). In this analysis, differences between populations were much larger than differences between the Inside and Outside locations.

The original analysis used an ANOVA on NMDS1. Since NMDS represents multivariate community data, PERMANOVA would be a more direct way to test overall community-composition differences. The ANOVA is kept here because it matches the original course analysis.
